---
title: "Exploring the HopFoehnTemp Data"
author: "Tino Schneidewind"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list= ls())
library(dplyr); library(tidyr); library(ggplot2) ; library(lubridate); library(plotly); library(viridis); library(gridExtra)
   

# notes for creating html document, paste this into Console
# rmarkdown::render("C:/Users/tinos/Documents/Master - Climate Science/3 - Master Thesis/master_thesis/vignettes/region_selection.Rmd", output_format = "html_document")
```


```{r data, echo=FALSE, warning=FALSE, fig.width=12}
# read Altdorf data with 5k buffer radius
altdorf = read.csv("C:/Users/tinos/Documents/Master - Climate Science/3 - Master Thesis/master_thesis/data/MedStat_aggregated/by_station/Altdorf_buffer_5000.csv")

# convert date to date format
altdorf$date <- as.Date(altdorf$date)


# aggregate by year
altdorf_yearly_sum = altdorf %>%
  mutate(year = as.Date(paste0(format(date, "%Y"), "-01-01"))) %>%  # Convert year to Date
  group_by(year) %>%
  summarize(across(c(all:uri, f_id, temp), sum, na.rm = TRUE))


altdorf_yearly_mean = altdorf %>%
  mutate(year = as.Date(paste0(format(date, "%Y"), "-01-01"))) %>%
  group_by(year) %>%
  summarize(across(c(all:uri, f_id, temp), mean, na.rm = TRUE))


# aggregate by month
altdorf_monthly_sum = altdorf %>%
  mutate(year = as.Date(paste0(format(date, "%m"), "-01-01"))) %>%
  group_by(year) %>%
  summarize(across(c(all:uri, f_id, temp), sum, na.rm = TRUE))

altdorf_monthly_mean = altdorf %>%
  mutate(year = as.Date(paste0(format(date, "%m"), "-01-01"))) %>%
  group_by(year) %>%
  summarize(across(c(all:uri, f_id, temp), mean, na.rm = TRUE))
```

```{r header, echo=FALSE}
# str(altdorf)
# summary(altdorf)
# head(altdorf)
```

Lets explore the data, and some descriptive statistics, its distribution, the time series etc. I will start of with **only the station Altdorf** as a placeholder for all individual stations. 

**How are hospitalizations distributed over the year?** Hospitalisations rates are highest in winter and lowest in summer. This is the case for men, women and all age groups. However, divided by diseases, this pattern seems to be driven by cardiovascular and respiratory diseases.

**How do yearly hospitalization develop over time?** Hospitalizations are increasing. There are rapid increases in 2008 and 2018. Men show higher hospitalization rates. Although the trend overall and in men could be due to demographics.

**What are the yearly hospitalizations per age and disease?** Here again, we see the rising trend which is strongest in the age group 15-64. 0-14 year olds even have a slight decreasing trend while all other age group slightly increase in hospitalizations over time. Divided by disease groups, cardiovascular diseases show the steepest increase with its curve resembling the overall hospitalization trend quite well. All other disease show a small trend but are hard to distinguish from another. Again, the interpretation of the observed trends is limited by the unknown demographic development.


```{r hosp, echo=FALSE}
# hospitalizations monthly
altdorf_monthly_mean_sex_long =  altdorf_monthly_mean[,c(1,2,8,9)] %>%
  pivot_longer(cols = c(all,mal,fem),
               names_to = "variable",         
               values_to = "value")  

p1 = ggplot(altdorf_monthly_mean_sex_long, aes(x = year, y = value, color = variable)) +
  geom_line() +
  theme_classic() +
  scale_color_viridis(discrete = TRUE) +
  labs(title = "average daily hosp per sex and month",
       x = "Month",
       y = "Hospitalization rate",
       color = "Variable")+
  theme(legend.position="right",  legend.margin=margin())



# hospitalizations monthly
altdorf_monthly_mean_sex_long =  altdorf_monthly_mean[,c(1,2,8,9)] %>%
  pivot_longer(cols = c(all,mal,fem),
               names_to = "variable",         
               values_to = "value")  

p1 = ggplot(altdorf_monthly_mean_sex_long, aes(x = year, y = value, color = variable)) +
  geom_line() +
  theme_classic() +
  scale_color_viridis(discrete = TRUE) +
  labs(title = "monthly daily hosp per sex",
       x = "",
       y = "Hospitalization rate",
       color = "")+
  theme(legend.position="right",  legend.margin=margin())


# monthly hosp per age category
altdorf_monthly_mean_age_long =  altdorf_monthly_mean[,c(1,3:7)] %>%
  pivot_longer(cols = c(a014y:a85plusy),
               names_to = "variable",         
               values_to = "value")  

p2 = ggplot(altdorf_monthly_mean_age_long, aes(x = year, y = value, color = variable)) +
  geom_line() +
  theme_classic() +
  scale_color_viridis(discrete = TRUE) +
  labs(title = "monthly daily hosp per age",
       x = "",
       y = "",
       color = "")+
  theme(legend.position="right",  legend.margin=margin())


# monthly hos per disease group
# monthly hosp per age category
altdorf_monthly_mean_disease_long =  altdorf_monthly_mean[,c(1,10:14)] %>%
  pivot_longer(cols = c(inf:uri),
               names_to = "variable",         
               values_to = "value")  

p3 = ggplot(altdorf_monthly_mean_disease_long, aes(x = year, y = value, color = variable)) +
  geom_line() +
  theme_classic() +
  scale_color_viridis(discrete = TRUE) +
  labs(title = "monthly daily hosp per disease",
       x = "Month",
       y = "",
       color = "")+
  theme(legend.position="right",  legend.margin=margin())

```


```{r hosp_2, echo=FALSE}
# hospitalizations total and per sex 
altdorf_yearly_sum_sex_long =  altdorf_yearly_sum[,c(1,2,8,9)] %>%
  pivot_longer(cols = c(all,mal,fem),
               names_to = "variable",         
               values_to = "value")  

o1 = ggplot(altdorf_yearly_sum_sex_long, aes(x = year, y = value, color = variable)) +
  geom_line() +
  theme_classic() +
  scale_color_viridis(discrete = TRUE) +
  labs(title = "yearly per sex",
       x = "Year",
       y = "Hospitalizations",
       color = "")+
  theme(legend.position="none",  legend.margin=margin())

# hospitalizations per disease and age category per year
altdorf_yearly_sum_age_long =  altdorf_yearly_sum[,c(1,3:7)] %>%
  pivot_longer(cols = c(a014y:a85plusy),
               names_to = "variable",         
               values_to = "value")  

o2 = ggplot(altdorf_yearly_sum_age_long, aes(x = year, y = value, color = variable)) +
  geom_line() +
  theme_classic() +
  scale_color_viridis(discrete = TRUE) +
  labs(title = "yearly per age group",
       x = "",
       y = "",
       color = "")+
  theme(legend.position="none", legend.margin=margin())


altdorf_yearly_sum_disease_long =  altdorf_yearly_sum[,c(1,10:14)] %>%
  pivot_longer(cols = c(inf : uri),
               names_to = "variable",         
               values_to = "value")  

o3 = ggplot(altdorf_yearly_sum_disease_long, aes(x = year, y = value, color = variable)) +
  geom_line() +
  theme_classic() +
 scale_color_viridis(discrete = TRUE) +
  labs(title = "yearly per disease group",
       x = "Year",
       y = "",
       color = "") +
  theme(legend.position="none", legend.margin=margin())

grid.arrange(p1,o1,p2,o2,p3,o3,ncol = 2)
```
The buffer datasets will be used to determine the trend within foehn.
Is there a trend for annual foehn counts?
Is there seasonality to foehn?

**Whats the annual path of temperature in Altdorf? Do we see an increasing trend of temperature?** As expected, it is warmest in summer with the highest monthly average in July just below 20 degrees C and coldest in winter just above 0 degrees. The yearly temperature trend has an increasing tendency. We see the hot years of 2003 and 2018 very clearly in Altdorf.

**How is temperature distributed?** Temperature has values between -10 and 30 degrees C average daily temperature. Temperature has no gaussian distribution but has in the median between the extremes a decrease in counts. 75% of all mean daily temperature values are between 5 and 15 degrees C.

```{r temp, echo=FALSE, fig.height=4}
# temperature monthly mean
altdorf_monthly_mean_temp =  altdorf_monthly_mean[,c(1,16)] %>%
  pivot_longer(cols = c(temp),
               names_to = "variable",         
               values_to = "value")  

o1 = ggplot(altdorf_monthly_mean_temp, aes(x = year, y = value)) +
  geom_line(color = "brown2") +
  theme_classic() +
  labs(title = "Monthly Temperature",
       x = "Month",
       y = "Temperature degrC") +
  theme(legend.position = "none", legend.margin = margin())


# temperature annual trend
altdorf_yearly_mean_temp =  altdorf_yearly_mean[,c(1,16)] %>%
  pivot_longer(cols = c(temp),
               names_to = "variable",         
               values_to = "value")  

o2 = ggplot(altdorf_yearly_mean_temp, aes(x = year, y = value)) +
  geom_line(color = "brown2") +
  theme_classic() +
  labs(title = "Yearly Temperature",
       x = "Year",
       y = "") +
  theme(legend.position = "none", legend.margin = margin())



o3 =ggplot(altdorf, aes(x = temp))+
  geom_histogram(color = "black", fill = "brown2", binwidth = 1, alpha = 0.7 )+
  theme_classic()+
  labs(title= "daily temperature distribution", x = "temperature in degrC")

o4 = ggplot(altdorf, aes(x= as.factor(station), y=temp ))+
  geom_boxplot( color = "black", fill = "brown2", alpha = 0.7)+
  theme_classic()+
  labs(title= "", x = "", y="temperature")

grid.arrange(o1,o2, o3, o4 ,ncol=2)
```



Do we have obvious correltaions in the dataset?
