---
title: "Foehn in Switzerland"
author: "Tino Schneidewind"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list= ls())
library(dplyr); library(tidyr); library(ggplot2) ; library(lubridate); library(plotly); library(viridis); library(gridExtra); library(corrplot); library(zoo)
   

# notes for creating html document, paste this into Console
# rmarkdown::render("C:/Users/tinos/Documents/Master - Climate Science/3 - Master Thesis/master_thesis/vignettes/Foehn.Rmd", output_format = "html_document")
```

### Foehn data investigation

In this analysis, all 8 stations are analyzed together and their foehn scores summed up for every day.  
```{r read data, echo=FALSE}
rm(list= ls())

# load data
data = read.csv("C:/Users/tinos/Documents/Master - Climate Science/3 - Master Thesis/master_thesis/data/MedStat_aggregated/hosp_buffer_5000.csv")

# convert date
data$date = as.Date(data$date)

# aggregate foehn by day: sum over all stations
foehn_d_sum <- data %>%
  group_by(date) %>%
  summarize(across(f_id, ~ sum(.x, na.rm = TRUE)))

# by month
foehn_m_mean <- foehn_d_sum %>%
  mutate(year_month =  paste0(format(date, "%Y-%m"),"-01" )) %>%
  group_by(year_month) %>%
  summarize(across(f_id, ~sum(.x, na.rm = TRUE))) %>%
  mutate(month =  as.numeric(format(as.Date(year_month), "%m")) )%>%
  group_by(month) %>%
  summarize(across(f_id, ~mean(.x, na.rm = TRUE))) 

# by year
foehn_y_sum <- foehn_d_sum %>%
  mutate(year =  as.numeric(format(date, "%Y"))) %>%
  group_by(year) %>%
  summarize(across(f_id, ~sum(.x, na.rm = TRUE)))

foehn_y_sum_summer <- foehn_d_sum %>%
  mutate(summer =  as.numeric(format(date, "%m"))) %>%
  filter(summer >5 & summer < 9) %>%
  mutate(year =  as.numeric(format(date, "%Y"))) %>%
  group_by(year) %>%
  summarize(across(f_id, ~sum(.x, na.rm = TRUE)))

foehn_y_sum_winter <- foehn_d_sum %>%
  mutate(winter =  as.numeric(format(date, "%m"))) %>%
  filter(winter < 3 | winter > 11) %>%
  mutate(year =  as.numeric(format(date, "%Y"))) %>%
  group_by(year) %>%
  summarize(across(f_id, ~sum(.x, na.rm = TRUE)))

foehn_y_sum_spring <- foehn_d_sum %>%
  mutate(winter =  as.numeric(format(date, "%m"))) %>%
  filter(winter > 2 & winter < 6) %>%
  mutate(year =  as.numeric(format(date, "%Y"))) %>%
  group_by(year) %>%
  summarize(across(f_id, ~sum(.x, na.rm = TRUE)))

foehn_y_sum_autumn <- foehn_d_sum %>%
  mutate(winter =  as.numeric(format(date, "%m"))) %>%
  filter(winter > 8 & winter < 12) %>%
  mutate(year =  as.numeric(format(date, "%Y"))) %>%
  group_by(year) %>%
  summarize(across(f_id, ~sum(.x, na.rm = TRUE)))

foehn_y_sum$winter <- foehn_y_sum_winter$f_id
foehn_y_sum$summer <- foehn_y_sum_summer$f_id
foehn_y_sum$spring <- foehn_y_sum_spring$f_id
foehn_y_sum$autumn <- foehn_y_sum_autumn$f_id

```

#### Seasonality of foehn in Switzerland:
- we have a spring peak
- very little foehn hours in summer
- average in autumn and winter

#### Distribution
- the frequency is antiproportional to the magnitude of the foehn day
```{r trends and seasonality, echo=FALSE, fig.height=3.5}
# monthly plot
p1 = ggplot(foehn_m_mean, aes(x = month, y = f_id)) +
  geom_line(color = "red3", lwd=1.2) +
  theme_classic() +
  scale_color_viridis(discrete = TRUE) +
  labs(title = "average monthly foehn score",
       x = "month",
       y = "foehn") +
  scale_x_continuous(breaks = unique(foehn_m_mean$month))


p2 = ggplot(foehn_d_sum[which(foehn_d_sum$f_id != 0),], aes(x = f_id))+
  geom_histogram(color = "black", fill = "red3", binwidth = 15, alpha = 0.7 )+
  theme_classic()+
  labs(title= "foehn score distr. without 0-foehn days", x = " foehn score")

grid.arrange(p1,p2,ncol=2)
```

#### Average Daily Foehn, excluding no foehn events, with 7 day running average
An average daily foehn score is: 79.6 (mean) // 18.51 (including 0), 59 (median) // 0 (including 0)
```{r average daily foehn intensity without no foehn events, echo=FALSE}
data$daymonth = format(data$date, "%m.%d")

foehn_d_mean <- data[data$f_id != 0,] %>%
  group_by(daymonth) %>%
  summarize(across(f_id, ~ mean(.x, na.rm = TRUE)))

foehn_d_median <- data[data$f_id != 0,]  %>%
  group_by(daymonth) %>%
  summarize(across(f_id, ~ median(.x, na.rm = TRUE)))

foehn_d_mean$date = paste0("2000.", foehn_d_mean$daymonth)
foehn_d_mean$date = as.Date(foehn_d_mean$date, format = "%Y.%m.%d")
foehn_d_median$date <- foehn_d_mean$date

plot(foehn_d_mean$date, foehn_d_mean$f_id, 
     xlab = "Date", ylab = "mean daily foehn", col = 2)
moving_mean <-c(NA,NA, NA,  rollmean(foehn_d_mean$f_id, k = 7), NA, NA,NA)
lines(foehn_d_mean$date, moving_mean, col = "black", lwd = 2)

# plot(foehn_d_mean$date, foehn_d_median$f_id,
#      xlab = "Date", ylab = "median daily foehn", col = 3)
moving_med <-c(NA,NA, NA,  rollmean(foehn_d_median$f_id, k = 7), NA, NA,NA)
lines(foehn_d_mean$date, moving_med, col = "green", lwd = 2)
legend("top", legend = c("7-day MA mean", "7-day MA median"), 
       col = c("black", "green"), lwd = 2, bty = "n")
```

#### Trends
- step increase in 2008 (due to different MedStat regions)
- otherwise visually decreasing for category: all
- spring is stationary
- all other seasons slightly decreasing

```{r trends, echo=FALSE, fig.height=3.5}

# trend plots
foehn_y_long = foehn_y_sum[,c(1:6)] %>%
  pivot_longer(cols = c(f_id: autumn),
               names_to = "variable", 
               values_to = "value")

ggplot(foehn_y_long, aes(x = year, y = value, color = variable)) +
  geom_line(lwd = 1) +
  theme_classic() +
  scale_color_viridis(discrete = TRUE, labels = c("all", "Dec-Feb", "Mar-May", "Jun-Aug", "Sep-Nov")) +
  labs(title = "Yearly foehn score",
       x = "year",
       y = "") +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 8),
        legend.position = "right")



foehn_y_long = foehn_y_sum[,c(1:6)] %>%
  pivot_longer(cols = c(f_id: autumn),
               names_to = "variable", 
               values_to = "value")
  
p3 =ggplot(foehn_y_long, aes(x = year, y = value, color = variable)) +
  geom_line(lwd = 1) +
  theme_classic() +
  scale_color_viridis(discrete = TRUE, labels = c("all", "Dec-Feb", "Mar-May", "Jun-Aug", "Sep-Nov")) +
  labs(title = "Yearly foehn score",
       x = "year",
       y = "") +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 8),
        legend.position = "right")

foehn_y_long = foehn_y_sum[1:10,c(1:6)] %>%
  pivot_longer(cols = c(f_id: autumn),
               names_to = "variable", 
               values_to = "value")
  
p3 =ggplot(foehn_y_long, aes(x = year, y = value, color = variable)) +
  geom_line(lwd = 1) +
  theme_classic() +
  scale_color_viridis(discrete = TRUE, labels = c("all", "Dec-Feb", "Mar-May", "Jun-Aug", "Sep-Nov")) +
  labs(title = "Yearly foehn score",
       x = "year",
       y = "") +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 8),
        legend.position = "right") 

# trend plots
foehn_y_long = foehn_y_sum[11:22,c(1:6)] %>%
  pivot_longer(cols = c(f_id: autumn),
               names_to = "variable", 
               values_to = "value")
  
p4 =ggplot(foehn_y_long, aes(x = year, y = value, color = variable)) +
  geom_line(lwd = 1) +
  theme_classic() +
  scale_color_viridis(discrete = TRUE, labels = c("all", "Dec-Feb", "Mar-May", "Jun-Aug", "Sep-Nov")) +
  labs(title = "Yearly foehn score",
       x = "year",
       y = "") +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 8),
        legend.position = "right") 

grid.arrange(p3,p4,ncol=2)


mean1 = mean(foehn_y_sum$f_id[1:10])
mean2 = mean(foehn_y_sum$f_id[11:22])
diff = mean2-mean1
foehn_y_sum$f_id_corr = foehn_y_sum$f_id
foehn_y_sum$f_id_corr[11:22] = foehn_y_sum$f_id_corr[11:22] - diff

foehn_y_long = foehn_y_sum[,c(1,7)] %>%
  pivot_longer(cols = c(f_id_corr),
               names_to = "variable", 
               values_to = "value")
  
ggplot(foehn_y_long, aes(x = year, y = value, color = variable)) +
  geom_line(lwd = 1) +
  theme_classic() +
  scale_color_viridis(discrete = TRUE, labels = c("all", "Dec-Feb", "Mar-May", "Jun-Aug", "Sep-Nov")) +
  labs(title = "Yearly foehn score corrected",
       x = "year",
       y = "") +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 8),
        legend.position = "right")




```

```{r pca, echo=FALSE}


#### PCA
# - all stations (alt, lug, mag, dav) have one dominating PC and then afterwards only homogeneous PCs
# - Biplots do not really show us anything but that the data is primarily of integer type
# - same for loading, looks to be  just the proportional weight of the observations and not a pattern.


# rm(list= ls())
# data = read.csv("C:/Users/tinos/Documents/Master - Climate Science/3 - Master Thesis/master_thesis/data/MedStat_aggregated/hosp_buffer_5000.csv")
# # for four stations
# alt <- data[data$station == "Altdorf",]
# lug <- data[data$station == "Lugano",]
# mag <- data[data$station == "Magadino",]
# dav <- data[data$station == "Davos",]
# 
# # pca
# alt_pca = prcomp(na.omit(alt[,c(3:15,17,18)]), center = TRUE, scale = TRUE)
# lug_pca = prcomp(na.omit(lug[,c(3:15,17,18)]), center = TRUE, scale = TRUE)
# mag_pca = prcomp(na.omit(mag[,c(3:15,17,18)]), center = TRUE, scale = TRUE)
# dav_pca = prcomp(na.omit(dav[,c(3:15,17,18)]), center = TRUE, scale = TRUE)
# 
# # screeplots
# par(mfrow=c(2,2))  
# plot(alt_pca, 
#      main = "Altdorf Screeplot", 
#      xlab = "Principal Component",
#      ylim = c(0,6))
# plot(lug_pca, 
#      main = "Lugano Screeplot", 
#      xlab = "Principal Component",
#      ylim = c(0,6))
# plot(mag_pca, 
#      main = "Magadino Screeplot", 
#      xlab = "Principal Component",
#      ylim = c(0,6))
# plot(dav_pca, 
#      main = "Davos Screeplot", 
#      xlab = "Principal Component",
#      ylim = c(0,6))
# 
# par(mfrow=c(1,2))
# 
# barplot(alt_pca$rotation[,1], las=2, main="Altdorf Loadings PC1") 
# # barplot(alt_pca$rotation[,2], las=2, main="Altdorf Loadings PC2")
# # barplot(alt_pca$rotation[,3], las=2, main="Altdorf Loadings PC3")
# # barplot(alt_pca$rotation[,4], las=2, main="Altdorf Loadings PC4")
# 
# 
# # biplot
# plot(alt_pca$x[,1], alt_pca$x[,2], type = "n", xlab = "Principal Component 1", ylab = "Principal Component 2", main = "Altdorf")
# # Add arrows
# arrows(0, 0, alt_pca$rotation[,1]*6, alt_pca$rotation[,2]*6, 
#        length = 0.1, angle = 15, col = "red")
# # Add dots
# text(alt_pca$x[,1], alt_pca$x[,2], labels = ".", col = "darkgray", cex = 3)
# # Add variable names
# text(alt_pca$rotation[,1]*6.5, alt_pca$rotation[,2]*6.5, 
#      labels = rownames(alt_pca$rotation), pos = 3, col = "black")

# loadings


```



```{r cluster, echo=FALSE}
# rm(list= ls())
# data = read.csv("C:/Users/tinos/Documents/Master - Climate Science/3 - Master Thesis/master_thesis/data/MedStat_aggregated/hosp_buffer_5000.csv")
# 
# X = data[1:200,c(3:15, 17,18)]
# #creating a dataset for each station
# stations = unique(data$station)
# 
# station_datasets <- list()
# 
# # Loop through each station
# for (i in stations) {
#   # Filter data for the current station
#   station_data <- filter(data, station == i)
#   
#   # Add the station data to the list
#   station_datasets[[i]] <- station_data[(nrow(station_data)-3999):nrow(station_data),c(3:15,17,18)]
# }
# 
# 
# station_datasets_scaled <- lapply(station_datasets, scale)
# 
# n= nrow(X)
# D_0 = matrix(Inf, nrow = n, ncol = n) 
# for (i in 1:(n-1)) {
#  for (j in (i+1):n) {
#  D_0[i,j] <- sqrt(sum((X[i,] - X[j,])^2)) # Euclidean distance
#  }
# }






```














